\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to express my sincere gratitude to all those who have supported me throughout this research journey.

First and foremost, I am deeply grateful to my supervisor, [Supervisor Name], for their invaluable guidance, patience, and expertise throughout this research. Their insights and constructive feedback have been instrumental in shaping this thesis.

I would also like to thank the members of my thesis committee, [Committee Member 1], [Committee Member 2], and [Committee Member 3], for their thoughtful comments and suggestions that have significantly improved the quality of this work.

My appreciation extends to my fellow graduate students and colleagues who provided stimulating discussions and moral support during the challenging phases of this research. Special thanks to [Colleague Names] for their collaboration and friendship.

I am grateful to [Institution/Organization] for providing the computational resources and datasets that made this research possible. The financial support from [Funding Source] is also acknowledged.

Finally, I would like to thank my family and friends for their unwavering support and encouragement throughout my academic journey. Their belief in me has been a constant source of motivation.

This work would not have been possible without the contributions of the entire research community working on concept drift and adaptive machine learning. I hope this thesis contributes meaningfully to this important field of study. 
